A Practical Optional Type System for Clojure

### 2012-10-29 dissertation/design-choices.tex
A bunch of corrections

We handle this by introducing a type \lstinline|(Array c t)|, which is an array has the Java component type \lstinline|c| 
array that can read and write values of type \lstinline|t|.
Internally, \lstinline|Array| types expand to \lstinline|(Array3 c w r)|,
which is an array has the Java component type \lstinline|c|,
that can write type \lstinline|w| (which is contravariant in \lstinline|Array|) and read type \lstinline|r| 
(which is covariant in \lstinline|Array|).
\lstinline|(Array c t)| expands to \lstinline|(Array3 c t t)|.

@@

Heterogeneous maps types in Typed Clojure are simple enough to avoid
the complications associated with concatenable records (see Section \ref{sec:recordtypes})
while still supporting many Clojure idioms.
They support adding new entries with keyword keys, removing entries,
and lookup of keyword keys with the type associated with each key in
a map tracked separately, similar to type systems for record types 
(see work by Harper and Pierce~\cite{HP91}).
When using these operations with non-keyword keys, heterogeneous maps
are treated like plain hash-maps.

>>>>

We handle this by introducing a type \lstinline|(Array c t)|, 
the type of the array that has the Java component type \lstinline|c| 
and that can read and write values of type \lstinline|t|.
Internally, \lstinline|Array| types expand to \lstinline|(Array3 c w r)|,
the type of the array that has the Java component type \lstinline|c|,
that can write values of type \lstinline|w| (which is contravariant in \lstinline|Array3|), 
and that can read values of type \lstinline|r| (which is covariant in \lstinline|Array3|).
Concretely, \lstinline|(Array c t)| expands to \lstinline|(Array3 c t t)|.

@@

Heterogeneous map types in Typed Clojure are simple enough to avoid
the complications associated with concatenable records (see Section \ref{sec:recordtypes})
while still supporting many Clojure idioms.
They support adding new entries with keyword keys, removing entries,
and lookup of keyword keys with the type associated with each key in
a map tracked separately, similar to type systems for record types 
(see work by Harper and Pierce~\cite{HP91}).
When using these operations with non-keyword keys, heterogeneous maps
are treated like plain hash-maps.

### 2012-10-29 dissertation/experiments.tex
A bunch of corrections

In this chapter work through several examples of using Typed Clojure and
how well the current prototype handles them. 
We intentially chose examples that could be challenging to type check
or were particularly useful to the everyday Clojure programmer.

@@

I ported a function from clojure.contrib.reflect\footnote{https://github.com/cemerick/pomegranate}, 
a Clojure library relying heavily on Java interoperability.
I chose \lstinline|call-method| for several reasons: it chains several Java calls together,
it uses primitive arrays, it exposes a wart of Clojure with \emph{Named} things,
and \lstinline|null| is a valid value in one place.

@@

The original function is modified slightly for readability and is presented in listing \ref{lst:callmethod}.

@@

Thankfully this function has up-to-date documentation, and from it we can derive an expected type
(listing \ref{lst:callmethodtype}).

@@

The final successfully type checked code is presented in listing \ref{lst:callmethodfinal}.

@@

A direct typing of the identity monad would look like listing \ref{lst:identitymtype}.
This signature assigns the expected type for the var \lstinline|identity-m| 
to be a heterogeneous map type with the minimum entries for monads: \lstinline|:m-bind|
and \lstinline|:m-result|. The syntax for heterogeneous map types requires a \lstinline|'| prefix,
and any number of key-value pairs are given between curly braces. Heterogeneous map keys must be keywords,
so the syntax is made more convenient by omitting the usually required \lstinline|'| prefix
for keyword types.

@@

We can now express the type of the identity monad more abstractly (listing \ref{lst:identityctors}).
The purpose of the monad seems more apparent just from reading its type.
In this case, the fact that identity monad has no effect is reflected
by its type constructor returning exactly its argument.

@@

The final type checked definition for the identity monad is given in listing \ref{lst:identitydeftyped}.
In this case, just the monadic bind required annotation.

@@

Listing \ref{lst:severalmonads} shows several monad types. The alias \lstinline|MonadPlusZero|
is identical to \lstinline|Monad|, except both \lstinline|:m-zero| and \lstinline|:m-plus| are defined.
The definitions were type checked by adding type annotations in appropriate places, in a similar fashion
to the identity monad. A significant obstacle was discovered in while type checking monad definitions,
relating to filtering sequences with negative type information, discussed in section \ref{sec:filterneg}.

@@

The initial motivation for adding user defined type constructors to Typed Clojure
(as discussed in section \ref{sec:monaddef}) was to type check
monad transformer definitions.
Monad transformers are always parameterised by a monad type constructor, and should work for all
monad type constructors. Keeping the monad type constructor abstract allows us to determine
whether a monad transformer works \emph{for all} monads.

@@

The library had to be modified and simplified considerably to be able to type check.
Many existing conduits had troublesome variable-parameters, reminiscent of the
issues of assigning types to \lstinline|assoc| and \lstinline|partial|
(which is future work, see section \ref{sec:futurearity}).
In these cases, simplified versions of the conduits were created that 
took fixed arguments.

@@

Listing \ref{lst:conduittypes} shows the type constructor for conduit,
\lstinline|==>|. Notice the body of \lstinline|==>| includes a reference
to itself in a position where variance must be known in advance.
\lstinline|declare-alias-kind| is used to declare the kind of an alias.
Once an alias with an already declared kind is defined with \lstinline|def-alias|,
the defined kind must match the declared kind, otherwise a type error is thrown.
(The specifics of the types in listing \ref{lst:conduittypes} are not relevant to the rest of this discussion,
so they are not explained).

>>>>

In this chapter we work through several examples of using Typed Clojure and
gauge how well the current prototype handles them. 
We intentionally chose examples that could be challenging to type check,
or were particularly useful to the everyday Clojure programmer.

@@

I ported a function from clojure.contrib.reflect\footnote{https://github.com/richhickey/clojure-contrib},
a Clojure library relying heavily on Java interoperability.
I chose \lstinline|call-method| for several reasons: it chains several Java calls together,
it uses primitive arrays, it exposes a wart of Clojure with \emph{Named} things,
and \lstinline|null| is a valid value in one place.

@@

The original function is modified slightly for readability and is presented in Listing \ref{lst:callmethod}.

@@

Thankfully this function has up-to-date documentation, and from it we can derive an expected type
(Listing \ref{lst:callmethodtype}).

@@

The final successfully type checked code is presented in Listing \ref{lst:callmethodfinal}.

@@

A direct typing of the identity monad would look like Listing \ref{lst:identitymtype}.
This signature assigns the expected type for the var \lstinline|identity-m| 
to be a heterogeneous map type with the minimum entries for monads: \lstinline|:m-bind|
and \lstinline|:m-result|. The syntax for heterogeneous map types requires a \lstinline|'| prefix,
and any number of key-value pairs are given between curly braces. Heterogeneous map keys must be keywords,
so the syntax is made more convenient by omitting the usually required \lstinline|'| prefix
for keyword types.

@@

We can now express the type of the identity monad more abstractly (Listing \ref{lst:identityctors}).
The purpose of the monad seems more apparent just from reading its type.
In this case, the fact that identity monad has no effect is reflected
by its type constructor returning exactly its argument.

@@

The final type checked definition for the identity monad is given in Listing \ref{lst:identitydeftyped}.
In this case, just the monadic bind required annotation.

@@

Listing \ref{lst:severalmonads} shows several monad types. The alias \lstinline|MonadPlusZero|
is identical to \lstinline|Monad|, except both \lstinline|:m-zero| and \lstinline|:m-plus| are defined.
The definitions were type checked by adding type annotations in appropriate places, in a similar fashion
to the identity monad. A significant obstacle was discovered while type checking monad definitions,
related to filtering sequences with negative type information (discussed in Section \ref{sec:filterneg}).

@@

The initial motivation for adding user defined type constructors to Typed Clojure
(as discussed in Section \ref{sec:monaddef}) was to type check
monad transformer definitions.
Monad transformers are always parameterised by a monad type constructor, and should work for all
monad type constructors. Keeping the monad type constructor abstract allows us to determine
whether a monad transformer works \emph{for all} monads.

@@

The library had to be modified and simplified considerably to be able to type check.
Many existing conduits had troublesome variable-parameters, reminiscent of the
issues of assigning types to \lstinline|assoc| and \lstinline|partial|
(which is future work, see Section \ref{sec:futurearity}).
In these cases, simplified versions of the conduits were created that 
took fixed arguments.

@@

Listing \ref{lst:conduittypes} shows the type constructor for conduit,
\lstinline|==>|. Notice the body of \lstinline|==>| includes a reference
to itself in a position where variance must be known in advance.
\lstinline|declare-alias-kind| is used to declare the kind of an alias.
Once an alias with an already declared kind is defined with \lstinline|def-alias|,
the defined kind must match the declared kind, otherwise a type error is thrown.
(The specifics of the types in Listing \ref{lst:conduittypes} are not relevant to the rest of this discussion,
so they are not explained).

### 2012-10-29 dissertation/future-work.tex
A bunch of corrections

Multimethods in Clojure are an idiomatic and often used feature.
It would be essential for an optional static type system for Clojure to support
The major hurdle for type checking multimethod definitions is the
interaction between the multimethod dispatch mechanism and occurrence typing.

>>>>

Multimethods in Clojure are an idiomatic and often used feature.
It would be essential for an optional static type system for Clojure to support
multimethods if it is to be of practical use.
The major hurdle for type checking multimethod definitions is the
interaction between the multimethod dispatch mechanism and occurrence typing.

### 2012-10-29 dissertation/implementation.tex
A bunch of corrections

Typed Clojure was built for the Clojure programming language, whose compiler
and data structures are implementated in Java. Clojurescript is the first major Clojure dialect
to be written in Clojure, and it is likely future dialects of Clojure will follow this example. 
Where Clojure uses Java Classes and Interfaces, 
Clojurescript's compiler and data structures are written in terms of Clojure's
two core abstractions: protocols and datatypes.
It would desirable to port Typed Clojure to similar  Clojure dialects while keeping
the core of the implementation constant.

>>>>

Typed Clojure was built for the Clojure programming language, whose compiler
and data structures are implemented in Java. ClojureScript is the first major Clojure dialect
to be written in Clojure, and it is likely future dialects of Clojure will follow this example. 
Where Clojure uses Java Classes and Interfaces, 
ClojureScript's compiler and data structures are written in terms of Clojure's
two core abstractions: protocols and datatypes.
It would be desirable to port Typed Clojure to such Clojure dialects while keeping
the core of the implementation constant.

### 2012-10-29 dissertation/intro.tex
A bunch of corrections

In the last decade it has become increasingly common to enhance
dynamically typed languages with static type systems. 
This is idea not new (see work by Fagan, Cartwright, Aiken and Murphy~\cite{Fag91,CF91,AM91}), but recent attempts
are noteworthy for their broad success in matching many of the advantages of statically typed
languages (see work by Tobin-Hochstadt~\cite{Tob10}), 
notably due to the use of bidirectional checking (like work by Pierce and Turner~\cite{PT00}).
Instead of always attempting to infer types, this algorithm relies on programmer annotations
appearing in some natural places such as giving the type of each top-level function.

@@

Recently a number of languages have been created to or been modified to support aspects
of both static and dynamic typing.
Dart~\cite{Dart2012} (created by Google) is dynamically typed but offers a simple form of optional
static typing that specifically do not affect runtime semantics.
Typescript~\cite{TypeS2012} (created by Microsoft) adds an optional type system to Javascript,
a well-known dynamic language.
Typed Racket~\cite{TF08,Tob10} (by Tobin-Hochstadt et al.) goes further by offering safe interoperability
between typed and untyped modules by generating appropriate runtime assertions based
on expected static types.

@@

It is common in Clojure, like other dynamically-typed languages, to
encode implicit type invariants in conditional tests.
In listing \ref{lst:numvec2}, conditional tests are used
to refine the types of the bindings \lstinline|a| and \lstinline|b|.
Occurrence typing~\cite{TF10} is a technique useful for capturing these kinds
of type invariants (occurrence typing is discussed in further detail in Section \ref{sec:OccurrenceTyping}).

>>>>

In the last decade it has become increasingly common to enhance
dynamically typed languages with static type systems. 
This idea is not new (see work by Fagan, Cartwright, Aiken and Murphy~\cite{Fag91,CF91,AM91}), but recent attempts
are noteworthy for their broad success in matching many of the advantages of statically typed
languages (see work by Tobin-Hochstadt~\cite{Tob10}), 
notably due to the use of bidirectional checking (like work by Pierce and Turner~\cite{PT00}).
Instead of always attempting to infer types, this algorithm relies on programmer annotations
appearing in some natural places such as giving the type of each top-level function.

@@

Recently a number of languages have been created or modified to support aspects
of both static and dynamic typing.
Dart~\cite{Dart2012} (created by Google) is dynamically typed but offers a simple form of optional
static typing that specifically do not affect runtime semantics.
Typescript~\cite{TypeS2012} (created by Microsoft) adds an optional type system to Javascript,
a well-known dynamic language.
Typed Racket~\cite{TF08,Tob10} (by Tobin-Hochstadt et al.) goes further by offering safe interoperability
between typed and untyped modules by generating appropriate runtime assertions based
on expected static types.

@@

It is common in Clojure, like other dynamically typed languages, to
encode implicit type invariants in conditional tests.
In listing \ref{lst:numvec2}, conditional tests are used
to refine the types of the bindings \lstinline|a| and \lstinline|b|.
Occurrence typing~\cite{TF10} is a technique useful for capturing these kinds
of type invariants (occurrence typing is discussed in further detail in Section \ref{sec:OccurrenceTyping}).

### 2012-10-29 dissertation/lit-review.tex
A bunch of corrections

Intersection, union, and singleton types are best considered as extrinsic types.
Hayashi~\cite{Hay91} describes two type theories, ATT which includes
intersection, union, and singleton types, and ATTT which further extends ATT
to include refinement types, which can be classified as extrinsic types.
Typed Clojure can also be considered an extrinsic type system which supports
union, intersection, and singleton types, but in less sophisticated forms
than ATT and ATTT.
In particular, singleton types in ATTT a much more advanced than Typed Clojure,
which offers only singleton types for values like \lstinline|:a| and \lstinline|nil|
and, unlike ATTT, no way to parameterise over singletons.

@@

Other statically typed languages support Java interoperability with
similar goals, most completely, Scala.
Scala~\cite{OCD+} is tightly integrated with Java. It manages interactions
with Java's \emph{null} pointer by using an \emph{Option} type, which (if
used correctly) provides strong elimination for \emph{null}.
\emph{null} is still not expressible as a type in Scala, however.
Arrays are parameterised by a non-variant (or invariant) parameter,
which are checked using \emph{conservative approximation}~\cite{OSV08}. For example,
this checks that covariant parameters are only used in covariant positions.

@@

The pseudocode in listing \ref{lst:bidir} show both algorithms in action. Local type argument synthesis is able
to infer the type arguments to \lstinline|map| by observing the argument types of the first
argument to \lstinline|map| and the types of subsequent sequence arguments. Bidirectional checking
then \emph{synthesises} the resulting type of the expression by \emph{checking} each element
of \lstinline|[1 2]| is a subtype of \lstinline|a|, and each element of \lstinline|[2.1 3.2]| is a subtype of
\lstinline|b|. The result of the anonymous function argument is \emph{synthesised} from the type of
\lstinline|(+ a b)| as \lstinline|Float|. We now have sufficient information to 
synthesise the type of listing \ref{lst:bidir} to be \lstinline|(List Float)|.

@@

\emph{Colored} types contain extra contextual information, including the propagation direction
and missing parts of the type. They are generally useful
for describing ``information flow in polymorphic type systems with propagation-based type inference''~\cite{OZZ01}. 
Colored Local Type Inference is a candidate for

@@

Still, unbounded type variables have an advantage: their implementations are
are simpler in the presence of a \emph{Bottom} type. 
The constraint resolution algorithm for bounded variables
is more subtle, due to ``some surprising interactions between bounded quantifiers
and the \emph{Bot} type''~\cite{PT00}, described fully
by Pierce~\cite{Pie97}.

@@

Typed Racket is a statically typed sister language of Racket. It
attempts to preserve existing Racket idioms and aims type check
existing Racket code by simply adding top level type annotations~\cite{Tob10}.

>>>>

Intersection, union, and singleton types are best considered as extrinsic types.
Hayashi~\cite{Hay91} describes two type theories, ATT which includes
intersection, union, and singleton types, and ATTT which further extends ATT
to include refinement types, which can be classified as extrinsic types.
Typed Clojure can also be considered an extrinsic type system which supports
union, intersection, and singleton types, but in less sophisticated forms
than ATT and ATTT.
In particular, singleton types in ATTT are much more advanced than in Typed Clojure,
which offers only singleton types for values like \lstinline|:a| and \lstinline|nil|
and, unlike ATTT, no way to parameterise over singletons.

@@

Other statically typed languages have similar goals
to support Java interoperability, like, most completely, Scala.
Scala~\cite{OCD+} is tightly integrated with Java. It manages interactions
with Java's \emph{null} pointer by using an \emph{Option} type, which (if
used correctly) provides strong elimination for \emph{null}.
\emph{null} is still not expressible as a type in Scala, however.
Arrays are parameterised by a non-variant (or invariant) parameter,
which are checked using \emph{conservative approximation}~\cite{OSV08}. For example,
this checks that covariant parameters are only used in covariant positions.

@@

The pseudocode in listing \ref{lst:bidir} show both algorithms in action. Local type argument synthesis is able
to infer the type arguments to \lstinline|map| by observing the argument types of the first
argument to \lstinline|map| and the types of subsequent sequence arguments. Bidirectional checking
then \emph{synthesises} the resulting type of the expression by \emph{checking} each element
of \lstinline|[1 2]| is a subtype of \lstinline|a|, and each element of \lstinline|[2.1 3.2]| is a subtype of
\lstinline|b|. The result of the anonymous function argument is \emph{synthesised} from the type of
\lstinline|(+ a b)| as \lstinline|Float|. We now have sufficient information to 
synthesise the type of listing \ref{lst:bidir} to be \lstinline|(LazySeq Float)|.

@@

\emph{Colored} types contain extra contextual information, including the propagation direction
and missing parts of the type. They are generally useful
for describing ``information flow in polymorphic type systems with propagation-based type inference''~\cite{OZZ01}. 
Colored Local Type Inference is a candidate for future extensions 
to Typed Clojure's inference.

@@

Still, unbounded type variables have an advantage: their implementations are
simpler in the presence of a \emph{Bottom} type. 
The constraint resolution algorithm for bounded variables
is more subtle, due to ``some surprising interactions between bounded quantifiers
and the \emph{Bot} type''~\cite{PT00}, described fully
by Pierce~\cite{Pie97}.

@@

Typed Racket is a statically typed sister language of Racket. It
attempts to preserve existing Racket idioms and aims to type check
existing Racket code by simply adding top level type annotations~\cite{Tob10}.

### 2012-10-29 dissertation/paper.tex
A bunch of corrections

This dissertation describes an optional static typing system
that attempts to bring the advantages of static typing to the dynamically type language
Clojure, running on the Java Virtual Machine. 
By reusing many of the design choices made in similar projects
like Typed Racket, we are able to design and implement a prototype type system
\emph{Typed Clojure} that can statically check many Clojure idioms.

>>>>

This dissertation describes an optional static typing system
that attempts to bring the advantages of static typing to the dynamically typed language
Clojure, running on the Java Virtual Machine. 
By reusing many of the design choices made in similar projects
like Typed Racket, we are able to design and implement a prototype type system
\emph{Typed Clojure} that can statically check many Clojure idioms.
